Covariate analysis of ROSMAP Reprocessed RNASeq

### 2016-07-20 code/Rmd/MSSM_RNASeq_Reprocessed.Rmd

tmp <- paste('Using following covariates in the final model:', modelStr)
```
`r tmp`
### Sanity check
```{r residualSigCovars.manual, fig.width=20, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
### Residual calculation
Calculate weighted residuals and add back "BrainRegion.Diagnosis" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("BrainRegion.Diagnosis", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```

@@

tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION)))
cols = WGCNA::labels2colors(COVARIATES.tmp);
tmp = gsub('\\..*','',cols)
WGCNA::plotDendroAndColors(tree, 
                           colors = tmp, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
```
```{r get.gene.symbols, cache=TRUE, eval=TRUE}
# Define biomart object
mart <- useMart('ENSEMBL_MART_ENSEMBL', dataset = "hsapiens_gene_ensembl", host = 'dec2015.archive.ensembl.org')

@@

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                       mart = mart)
RESIDUAL.GENE_EXPRESSION <- rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, "Gene.ID") %>%
  left_join(backgroundGenes) %>%
  left_join(Ensemble2HGNC) %>% 
  dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
VOOM.ADJUSTED.GENE_EXPRESSION$E <- rownameToFirstColumn(VOOM.ADJUSTED.GENE_EXPRESSION$E, "Gene.ID") %>%
  left_join(backgroundGenes) %>%
  left_join(Ensemble2HGNC) %>% 
  dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```
```{r summary, results='asis', echo=FALSE}
writeLines('Number of samples in each category are')
tmp = summary(COVARIATES$BrainRegion.Diagnosis) %>%
  rownameToFirstColumn('BrainRegion.Diagnosis') %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion','Diagnosis'), sep = '\\.') %>%
  tidyr::spread(Diagnosis, DF)
kable(tmp)
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=12, fig.width=25}
# Fit contrast
contrast = makeContrasts(contrasts = sapply(unique(METADATA$BrainRegion), function(x){
  apply(combn(paste0('BrainRegion.Diagnosis',x,'.',unique(METADATA$Diagnosis)),2), 2, paste, collapse = '-')
}) %>%
  as.character(), 
levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
FIT = contrasts.fit(VOOM.ADJUSTED.FIT, contrasts=contrast)
FIT = eBayes(FIT)

>>>>

tmp <- paste('Using following covariates in the final model:', modelStr)
```
`r tmp`
### Normalisation (manual design)
Adding AOD as a covariate to the identified covariates (for cross brain bank comparison)
```{r manualNorm}
# Primary variable of interest
postAdjustCovars = c(postAdjustCovars, 'Batch', 'PMI', 'AOD') %>% unique

writeLines(paste('Using following covariates in the model:',
                 paste(postAdjustCovars, collapse=', '),
                 'as fixed effects'))

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                              COVARIATES, 'all adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
```
### Sanity check
```{r residualSigCovars.manual, fig.width=20, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
### Residual calculation
Calculate weighted residuals and add back "BrainRegion.Diagnosis" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("BrainRegion.Diagnosis", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```

@@

tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION)))
cols = WGCNA::labels2colors(COVARIATES.tmp);
tmp = gsub('\\..*','',cols)
WGCNA::plotDendroAndColors(tree, 
                           colors = tmp, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
```

```{r get.gene.symbols, cache=TRUE, eval=TRUE}
# Define biomart object
mart <- useMart('ENSEMBL_MART_ENSEMBL', dataset = "hsapiens_gene_ensembl", host = 'dec2015.archive.ensembl.org')

@@

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                       mart = mart)
RESIDUAL.GENE_EXPRESSION <- rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, "Gene.ID") %>%
  left_join(backgroundGenes) %>%
  left_join(Ensemble2HGNC) %>% 
  dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
VOOM.ADJUSTED.GENE_EXPRESSION$E <- rownameToFirstColumn(VOOM.ADJUSTED.GENE_EXPRESSION$E, "Gene.ID") %>%
  left_join(backgroundGenes) %>%
  left_join(Ensemble2HGNC) %>% 
  dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```

```{r summary, results='asis', echo=FALSE}
writeLines('Number of samples in each category are')
tmp = summary(COVARIATES$BrainRegion.Diagnosis) %>%
  rownameToFirstColumn('BrainRegion.Diagnosis') %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion','Diagnosis'), sep = '\\.') %>%
  tidyr::spread(Diagnosis, DF)
kable(tmp)
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=12, fig.width=25}
# Fit contrast
contrast = makeContrasts(contrasts = sapply(unique(METADATA$BrainRegion), function(x){
  apply(combn(paste0('BrainRegion.Diagnosis',x,'.',unique(METADATA$Diagnosis)),2), 2, paste, collapse = '-')
}) %>%
  as.character(), 
levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
FIT = contrasts.fit(VOOM.ADJUSTED.FIT, contrasts=contrast)
FIT = eBayes(FIT)

### 2016-07-20 code/Rmd/Mayo_RNASeq_Reprocessed_AdjustPMI.Rmd

tmp = DE %>% 
  dplyr::filter(adj.P.Val <= 0.05) %>%
  group_by(Comparison) %>%
  dplyr::summarize(nDiffExpGenes = length(unique(Gene.ID)))
kable(tmp)

>>>>

tmp = DE %>% 
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::mutate(D = logFC/abs(logFC)) %>%
  group_by(Comparison, D) %>%
  dplyr::summarize(nDiffExpGenes = length(unique(Gene.ID))) %>%
  spread(D, nDiffExpGenes)
kable(tmp)

### 2016-07-20 code/Rmd/Mayo_RNASeq_SVA_Reprocessed.Rmd

# library(doParallel)
# library(foreach)
# 
# cl = makeCluster(14)
# registerDoParallel(cl)

@@

# AgeAtDeath
p[[2]] = ggplot(METADATA, aes(x = Diagnosis, y = as.numeric(AgeAtDeath), fill = Tissue)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('AgeAtDeath') + theme(legend.position = 'top')

multiplot(plotlist = p, cols = 2)

@@

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis")
ContCovariates <- c("RIN", "RIN2", "AgeAtDeath")

@@

# Find significant covariates
adjustCovars = preAdjustedSigCovars$significantCovars
```
Significant covariates to adjust at FDR 0.1 are `r paste(gsub('_','\\\\_',adjustCovars), collapse= ',')`
```{r preAdjustedSigCovars.NULL.ALL, fig.width=25, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (iterative)
Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix.
```{r iterative.adjusted.voom.normalisation, results='asis', echo=FALSE}
postAdjustCovars = c('BrainRegion.Diagnosis')
randomCovariates = 'Donor_ID'

# Assign residual covariates
residualSigCovars = preAdjustedSigCovars
significantCovars = residualSigCovars$significantCovars
covariatesEffects = residualSigCovars$Effects.significantCovars
covariatesEffects = covariatesEffects[setdiff(significantCovars, c(postAdjustCovars, randomCovariates))]

postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))

loopCount = 0 
while(length(significantCovars)!=0 && loopCount <= 100){
  writeLines(paste('Using following covariates in the model:', 
                   paste(postAdjustCovars, collapse=', '),
                   'as fixed effects'))

@@

  # Residual covariates to choose from
  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars, randomCovariates))

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                                COVARIATES,
                                                'all adjusted design(voom-normalized)',
                                                isKeyPlot=TRUE)

  # Add postadjusted covariates (if any)
  significantCovars = residualSigCovars$significantCovars
  covariatesEffects = residualSigCovars$Effects.significantCovars
  covariatesEffects = covariatesEffects[setdiff(significantCovars,c(postAdjustCovars, randomCovariates))]

  postAdjustCovars = c(postAdjustCovars,names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}
modelStr <-paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=','), 'as fixed effects')

tmp <- paste('Using following covariates in the final model:', modelStr, 'and Donor_ID as random effects')
```
`r tmp`
### Sanity Check
Residualise significant covariates using a linear model and find correlation between PCA of residuals with covariates
```{r residualSigCovars.ALL, fig.width=25, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
### Estimate surrogate variables using SVA package
```{r sva}
MODEL0 = getDesignMatrix(COVARIATES[,setdiff(postAdjustCovars, "BrainRegion.Diagnosis"),drop=F],Intercept = F)$design # Get null model by removing variable of interest
MODEL1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)$design

## Calculate the number of surrogate variables
# Get number of surrogate variables (for details of the algorithm refer http://biorxiv.org/content/biorxiv/suppl/2016/05/09/052209.DC1/052209-1.pdf)
SINGULAR.VAL.DECOM <- svd(RESIDUAL.GENE_EXPRESSION)
NUM.DEG.FREEDOM = dim(RESIDUAL.GENE_EXPRESSION)[2] - dim(MODEL1)[2] + 1
SING.VALS <- SINGULAR.VAL.DECOM$d^2/sum(SINGULAR.VAL.DECOM$d)
SING.VALS = SING.VALS[1:NUM.DEG.FREEDOM]

# Permute and calculate the permuted singular values
PERM.SING.VALS <- foreach (i = 1:100,
         .combine = rbind,
         .packages = c('limma','CovariateAnalysis'),
         .export = c('EXPR','MODEL1')) %dopar% {

           # Permute expression counts
           PERM.COUNTS <- t(apply(EXPR$filteredExprMatrix, 1, sample, replace = FALSE))

           # Estimate voom weights
           VOOM.ADJUSTED.PERM.GENE_EXPRESSION = voom(PERM.COUNTS, design=MODEL1, plot=F)

           # Fit linear model using new weights and new design
           VOOM.ADJUSTED.PERM.FIT = lmFit(VOOM.ADJUSTED.PERM.GENE_EXPRESSION)

           # Residuals after normalisation
           RESIDUAL.PERM.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.PERM.FIT,VOOM.ADJUSTED.PERM.GENE_EXPRESSION$E)

           PERM.SINGULAR.VAL.DECOM <- svd(RESIDUAL.PERM.GENE_EXPRESSION)
           PERM.SING.VALS <- PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2/sum(PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2)
        }
num.sv1 = sum(SING.VALS > t(rowMeans(t(PERM.SING.VALS)))[1,1])

num.sv2 = num.sv(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, method = 'be', seed = 123456)
NUM.SV = min(num.sv1, num.sv2)

@@

names(clust) = sampleTrees$labels
clust = clust[rownames(COVARIATES)]
```
```{r decompse.normalise.data4, fig.height=10, fig.width=25, results='asis'}
# Tree based clustering
COVARIATES.tmp = COVARIATES
COVARIATES.tmp[, FactorCovariates] = sapply(COVARIATES.tmp[,FactorCovariates], as.numeric)
plotDendroAndColors(sampleTrees, 
                    colors = labels2colors(cbind(data.frame(clust = clust), COVARIATES.tmp)), 
                    abHeight = 0.80, 
                    main = "Sample dendrogram",
                    groupLabels = colnames(COVARIATES))

tmp = paste('Samples in cluster 0 are', paste(sampleTrees$labels[clust == 0], collapse = ','))
tmp = base::gsub('_','\\\\_',tmp)
writeLines(tmp)
tmp = paste('Samples in cluster 3 are', paste(sampleTrees$labels[clust == 3], collapse = ','))
tmp = base::gsub('_','\\\\_',tmp)
writeLines(tmp)
tmp = paste('Samples in cluster 4 are', paste(sampleTrees$labels[clust == 4], collapse = ','))
tmp = base::gsub('_','\\\\_',tmp)
writeLines(tmp)
```

@@

VOOM.ADJUSTED.GENE_EXPRESSION$E <- rownameToFirstColumn(VOOM.ADJUSTED.GENE_EXPRESSION$E, 'Gene.ID') %>%
  left_join(backgroundGenes) %>% left_join(Ensemble2HGNC) %>%
  dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=15, fig.width=25}
# Fit contrast
contrast = makeContrasts(contrasts=c(
  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.AD",
  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.Pathologic_Aging",
  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.PSP",
  "BrainRegion.DiagnosisCER.Pathologic_Aging-BrainRegion.DiagnosisCER.AD",
  "BrainRegion.DiagnosisCER.PSP-BrainRegion.DiagnosisCER.AD",
  "BrainRegion.DiagnosisCER.Pathologic_Aging-BrainRegion.DiagnosisCER.PSP",
  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.AD",
  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.Pathologic_Aging",
  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.PSP",
  "BrainRegion.DiagnosisTCX.Pathologic_Aging-BrainRegion.DiagnosisTCX.AD",
  "BrainRegion.DiagnosisTCX.PSP-BrainRegion.DiagnosisTCX.AD",
  "BrainRegion.DiagnosisTCX.Pathologic_Aging-BrainRegion.DiagnosisTCX.PSP"), 
                         levels = colnames(FIT$coefficients))
FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

@@

>>>>

library(glmnet)

library(doParallel)
library(foreach)

cl = makeCluster(14)
registerDoParallel(cl)

@@

# AgeAtDeath
p[[2]] = ggplot(METADATA, aes(x = Diagnosis, y = as.numeric(AgeAtDeath), fill = Tissue)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('AgeAtDeath') + theme(legend.position = 'top')
# PMI
p[[3]] = ggplot(METADATA, aes(x = Diagnosis, y = as.numeric(PMI), fill = Tissue)) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PMI') + theme(legend.position = 'top')

multiplot(plotlist = p, cols = 3)

@@

### Covariate clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c("Donor_ID", "Source", "Gender", "FLOWCELL", "BrainRegion.Diagnosis")
ContCovariates <- c("RIN", "RIN2", "AgeAtDeath", "PMI")

@@

# Find significant covariates
adjustCovars = preAdjustedSigCovars$significantCovars
```
Significant covariates to adjust at FDR 0.1 are `r paste(gsub('_','\\\\_',adjustCovars), collapse= ',')`
```{r preAdjustedSigCovars.NULL.ALL, fig.width=25, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (manual design)
RIN and PMI. Since there is a strong association of case control status with Source and flowcell information we are manually including them. Also, to be consistent with other brain banks we are including AgeAtDeath as a covariate.
```{r manual.adjusted.voom.normalisation, results='asis', echo=FALSE}
postAdjustCovars = c('BrainRegion.Diagnosis', 'RIN', 'PMI', 'Source', 'FLOWCELL', 'AgeAtDeath')
randomCovariates = 'Donor_ID'

writeLines(paste('Using following covariates in the model:', 
                 paste(postAdjustCovars, collapse=', '),
                 'as fixed effects and',
                 paste(randomCovariates, collapse=', '),
                 'random effect'))

@@

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                              COVARIATES,
                                              'all adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
```
### Sanity Check
Residualise significant covariates using a linear model and find correlation between PCA of residuals with covariates
```{r residualSigCovars.ALL, fig.width=25, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
### Estimate surrogate variables using SVA package
```{r sva}
MODEL0 = getDesignMatrix(COVARIATES[,setdiff(postAdjustCovars, "BrainRegion.Diagnosis"),drop=F],Intercept = F)$design # Get null model by removing variable of interest
MODEL1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)$design
MODEL1 = MODEL1[,linColumnFinder(MODEL1)$indepCols]

# ## Calculate the number of surrogate variables
# # Get number of surrogate variables (for details of the algorithm refer http://biorxiv.org/content/biorxiv/suppl/2016/05/09/052209.DC1/052209-1.pdf)
# SINGULAR.VAL.DECOM <- svd(RESIDUAL.GENE_EXPRESSION)
# NUM.DEG.FREEDOM = dim(RESIDUAL.GENE_EXPRESSION)[2] - dim(MODEL1)[2] + 1
# SING.VALS <- SINGULAR.VAL.DECOM$d^2/sum(SINGULAR.VAL.DECOM$d)
# SING.VALS = SING.VALS[1:NUM.DEG.FREEDOM]
# 
# # Permute and calculate the permuted singular values
# PERM.SING.VALS <- foreach (i = 1:100,
#          .combine = rbind,
#          .packages = c('limma','CovariateAnalysis'),
#          .export = c('EXPR','MODEL1')) %dopar% {
#            
#            # Permute expression counts
#            PERM.COUNTS <- t(apply(EXPR$filteredExprMatrix, 1, sample, replace = FALSE))
#            
#            # Estimate voom weights
#            VOOM.ADJUSTED.PERM.GENE_EXPRESSION = voom(PERM.COUNTS, design=MODEL1, plot=F)
#            
#            # Fit linear model using new weights and new design
#            VOOM.ADJUSTED.PERM.FIT = lmFit(VOOM.ADJUSTED.PERM.GENE_EXPRESSION)
#            
#            # Residuals after normalisation
#            RESIDUAL.PERM.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.PERM.FIT,VOOM.ADJUSTED.PERM.GENE_EXPRESSION$E)
#            
#            PERM.SINGULAR.VAL.DECOM <- svd(RESIDUAL.PERM.GENE_EXPRESSION)
#            PERM.SING.VALS <- PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2/sum(PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2)
#         }
# num.sv1 = sum(SING.VALS > t(rowMeans(t(PERM.SING.VALS)))[1,1])

num.sv2 = num.sv(VOOM.ADJUSTED.GENE_EXPRESSION$E, MODEL1, method = 'be', seed = 123456)
NUM.SV = num.sv2#min(num.sv1, num.sv2)

@@

names(clust) = sampleTrees$labels
clust = clust[rownames(COVARIATES)]
```
```{r decompse.normalise.data4, fig.height=10, fig.width=25, results='asis'}
# Tree based clustering
COVARIATES.tmp = COVARIATES
COVARIATES.tmp[, FactorCovariates] = sapply(COVARIATES.tmp[,FactorCovariates], as.numeric)
col.mat = labels2colors(cbind(tree,COVARIATES.tmp))
col.mat = gsub('\\.1', '', col.mat)

plotDendroAndColors(sampleTrees, 
                    colors = col.mat,
                    abHeight = 0.80, 
                    main = "Sample dendrogram",
                    groupLabels = colnames(COVARIATES))
```

@@

VOOM.ADJUSTED.GENE_EXPRESSION$E <- rownameToFirstColumn(VOOM.ADJUSTED.GENE_EXPRESSION$E, 'Gene.ID') %>%
  left_join(backgroundGenes) %>% left_join(Ensemble2HGNC) %>%
  dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=10, fig.width=20}
# Fit contrast
contrast = makeContrasts(contrasts=c(
  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.AD",
  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.Pathologic_Aging",
  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.PSP",
  "BrainRegion.DiagnosisCER.Pathologic_Aging-BrainRegion.DiagnosisCER.AD",
  "BrainRegion.DiagnosisCER.PSP-BrainRegion.DiagnosisCER.AD",
  "BrainRegion.DiagnosisCER.Pathologic_Aging-BrainRegion.DiagnosisCER.PSP",
  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.AD",
  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.Pathologic_Aging",
  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.PSP",
  "BrainRegion.DiagnosisTCX.Pathologic_Aging-BrainRegion.DiagnosisTCX.AD",
  "BrainRegion.DiagnosisTCX.PSP-BrainRegion.DiagnosisTCX.AD",
  "BrainRegion.DiagnosisTCX.Pathologic_Aging-BrainRegion.DiagnosisTCX.PSP"), 
                         levels = colnames(FIT$coefficients))
FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

@@

# Design Matrix
DM = VOOM.ADJUSTED.GENE_EXPRESSION$design
DM = DM[,!(colnames(DM) %in% c('PMI'))]
write.table(DM, file='MAYOdesign.tsv', sep = '\t', row.names=F, quote=F)
DM_OBJ = File('MAYOdesign.tsv', 
                  name = 'Design Matrix',
                  parentId = CODE$properties$id)
DM_OBJ = synStore(DM_OBJ,
                      used = ALL_USED_IDs, 
                      activityName = activityName, 
                      executed = thisFile, 
                      activityDescription = activityDescription)

### 2016-07-20 code/Rmd/ROSMAP_RNASeq_Reprocessed_NonConditionedCogdx.Rmd

knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_Reprocessed.Rmd",
                                 parentId ="syn5570291",
                                 entityName = 'ROSMAP RNASeq Reprocessed Counts (Based on cogdx)')
```

@@

@@

@@

thisFileName <- 'ROSMAP_RNASeq_Reprocessed.Rmd'

@@

### Normalisation (iterative design)
Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix. Here "cogdx" is chosen as the primary variable of interest
```{r iterativeNorm}
# Primary variable of interest
postAdjustCovars = 'cogdx'

@@

  # Post adjusted design matrix
  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

@@

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                       mart = mart)

# Add gene names
RESIDUAL.GENE_EXPRESSION <- RESIDUAL.GENE_EXPRESSION %>%
  rownameToFirstColumn('Gene.ID') %>%
  left_join(backgroundGenes) %>%
  left_join(Ensemble2HGNC) %>%
  dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))

VOOM.ADJUSTED.GENE_EXPRESSION$E <- VOOM.ADJUSTED.GENE_EXPRESSION$E %>%
  rownameToFirstColumn('Gene.ID') %>%
  left_join(backgroundGenes) %>%
  left_join(Ensemble2HGNC) %>%
  dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=8, fig.width=16}
# Fit contrast
contrast = makeContrasts(contrasts=c("cogdx1-cogdx2",
                                     "cogdx1-cogdx4",
                                     "cogdx2-cogdx4"), 
                         levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
FIT.CONTR = contrasts.fit(VOOM.ADJUSTED.FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

@@

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC)) + geom_point() + xlim(c(-1,1))
p = p + geom_hline(yintercept = -log10(0.05), color = 'red') 
p = p + geom_vline(xintercept = log2(1.2), color = 'red') 
p = p + geom_vline(xintercept = -log2(1.2), color = 'red') 
p = p + facet_grid(.~Comparison, scales = 'fixed')
p
```
### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = "ROSMAP RNASeq Reprocessed Counts (Based on cogdx)", parentId = parentId)
CODE <- synStore(CODE)

>>>>

knit2synapse::knitToFolderEntity(file = "./ROSMAP_RNASeq_Reprocessed_NonConditionedCogdx.Rmd",
                                 parentId ="syn5570291",
                                 entityName = 'ROSMAP RNASeq Reprocessed Counts 2 (Based on cogdx)')
```

@@

library(ggdendro)
library(RColorBrewer)

@@

devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")

@@

thisFileName <- 'ROSMAP_RNASeq_Reprocessed_NonConditionedCogdx.Rmd'

@@

### Normalisation (iterative design)
Since many covariates are correlated, re-normalising COUNTS with an iterative design matrix. Here "cogdx" is chosen as the primary variable of interest
```{r iterativeNorm}
# Primary variable of interest
postAdjustCovars = NULL

@@

  # Post adjusted design matrix
  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols, drop=F]

@@

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                       mart = mart)
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=8, fig.width=16}
# Post adjusted design matrix
DM1 = getDesignMatrix(COVARIATES[,'cogdx',drop=F],Intercept = F)
DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols, drop=F]

# Fit linear model using residuals and new design matrix
FIT = lmFit(RESIDUAL.GENE_EXPRESSION, DM1$design)

# Fit contrast
contrast = makeContrasts(contrasts=c("cogdx1-cogdx2",
                                     "cogdx1-cogdx4",
                                     "cogdx2-cogdx4"), 
                         levels = colnames(FIT$coefficients))
FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

@@

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC)) + geom_point() + xlim(c(-1,1))
p = p + geom_hline(yintercept = -log10(0.05), color = 'red') 
p = p + geom_vline(xintercept = log2(1.2), color = 'red') 
p = p + geom_vline(xintercept = -log2(1.2), color = 'red') 
p = p + facet_grid(.~Comparison, scales = 'fixed')
p
```
### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Add gene names
RESIDUAL.GENE_EXPRESSION <- RESIDUAL.GENE_EXPRESSION %>%
  rownameToFirstColumn('Gene.ID') %>%
  left_join(backgroundGenes) %>%
  left_join(Ensemble2HGNC) %>%
  dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))

VOOM.ADJUSTED.GENE_EXPRESSION$E <- VOOM.ADJUSTED.GENE_EXPRESSION$E %>%
  rownameToFirstColumn('Gene.ID') %>%
  left_join(backgroundGenes) %>%
  left_join(Ensemble2HGNC) %>%
  dplyr::select(Gene.ID, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))

# Code
CODE <- Folder(name = "ROSMAP RNASeq Reprocessed Counts 2 (Based on cogdx)", parentId = parentId)
CODE <- synStore(CODE)
